Prolonged exposure to pain circumstances can have many side-effects on the life of a fetus and cause negative developmental consequences. Thus, pain assessment and management is made necessary to identify these scenarios early on. Even though numerous pain scales exist to help assess pain in neonates, until recently, no such method existed for detecting pain in fetuses. Based on these scales, some research has been developed to automatically assess pain through the means of analyzing images with computational help. Still, no such work had been developed for fetuses as well.

In this scenario, we propose the use of deep convolutional neural networks to construct a learning model capable of automatically detecting the presence of pain in fetuses. We do so through the evaluation of their facial expressions in images collected from 4-D ultrasound machines. By taking advantage of transfer learning, we used a network pre-trained on the task of face recognition, and confirmed that transferring from a similar task performed better than if made from a general-purpose dataset. 

We have evaluated our model on images extracted from 13 video recordings of fetuses undergoing painful and non-painful stimulus and achieved an accuracy of 78.7\% on the task of discriminating images of pain from those in a non-painful control group. Our results demonstrate the effectiveness of applying such methods with fetal images, and above all, show that it is possible to develop a model for automatically detecting pain in fetuses.

\keywords{Deep Learning, Transfer Learning, Fetal Pain, Automatic Pain Assessment}
